What We Eval in the Shadows

### 2021-08-10 paper/main.tex
Add more discussion in 5.1

\paragraph{Discussion}

\Eval is used in CRAN packages. Nearly every fourth package contains at least
one \eval site. However, most packages that use \eval do that modestly. Half of
them include less than three \eval sites. Packages that rely heavily on \eval
are mostly the ones that provide statistical modeling and simulation functions.
At runtime, over a third of extracted programs triggered calls to \eval. Half
of the \packageAllcallsRnd recorded calls come from a single \eval site, but
for the rest, most of the packages have low \eval call frequency.

@@

\paragraph{Discussion}

It is interesting how many different forms any given site sees. The more forms,
the harder it is to characterize behavior at that site. Luckily,
\packageNbOneMinimizedPercent of sites only see a single form. In other words,
a compiler could correctly predict the \emph{shape} of an expression by looking
just at the first call in \packageNbOneMinimizedPercent of the sites. Few sites
are highly polymorphic (8 or more different forms). These include the pipe
operator of the \c{magrittr} package, used to compose functions, \eg instead of
\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}.

The variety of uses of \eval is evidenced by the number of different forms. In
comparison, JavaScript \eval usage was more straightforward and more
predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with
only one form. On the other hand, there are \packageNbSimpleMinimizedOne sites
that only receive one of the simple forms (\ie\xspace \c{X}, \c{V}, \c{\$} or
\c{<-}). Additional \packageNbSimpleMinimizedMore sites takes multiple forms,
but all are these simple ones. Among these sites there are many cases where
\eval could be replaced by less powerful constructs such as \c{get} to lookup a
name in an environment, \c{assign} to set a value to a name in an environment
or \c{do.call} to reflectively call a function.

@@

\paragraph{Discussion}
The data presented in this section is what one could expect. For {\it
  functions}, they represent the main user case for \eval. Thus \eval is really
extending the behavior of that function by executing a dynamically selected
piece of code in the function's scope. Typically, variables are read (mostly),
written (less), but there are also some cases where new variables are injected
or existing variables are deleted. Usually, this happens in the current
function, but also in frames arbitrarily far up the call stack. One data point
we lack is how the target environment was obtained. The expected case is that
the body of a promise was modified and the resulting expression was evaluated in
the same environment the promise originated from. Likely less frequent are cases
where \eval is provided the results of programmatically selecting some call
stack. For {\it synthetics}, their relatively high frequency correlates to use
cases where one wants to evaluate an expression in either a restricted
environment or to use a data structure as an environment. For {\it global}, its
relatively high frequency is likely an artifact of how the code of our corpus is
is run. For {\it packages}, only \packageNbPackageNamespaceEnvSites sites have
this kind. This is probably for the best as mutating the bindings of
a loaded package is improper and R tries to make it difficult.

@@

\paragraph{Discussion}
The origin data suggests that constructing expressions from strings is a
minority of the use cases. Instead, the constructed category shows that most
\evals comes from code that was processed by the compiler, and may be slightly
modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those. Another use of \eval is to directly build symbols to evaluate them in another environment than the default one.

>>>>

\mypara{Discussion.}
\Eval is used in CRAN packages. Nearly every fourth package contains at least
one \eval site. However, most packages that use \eval do that modestly. Half of
them include less than three \eval sites. Packages that rely heavily on \eval
are mostly the ones that provide statistical modeling and simulation functions.
At runtime, over a third of extracted programs triggered calls to \eval. Half
of the \packageAllcallsRnd recorded calls come from a single \eval site, but
for the rest, most of the packages have low \eval call frequency.

@@

\mypara{Discussion.} 
It is interesting how many different forms any given site sees. The more forms,
the harder it is to characterize behavior at that site. Luckily,
\packageNbOneMinimizedPercent of sites only see a single form. In other words,
a compiler could correctly predict the \emph{shape} of an expression by looking
just at the first call in \packageNbOneMinimizedPercent of the sites. Few sites
are highly polymorphic (8 or more different forms). These include the pipe
operator of the \c{magrittr} package, used to compose functions, \eg instead of
\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}.

The variety of uses of \eval is evidenced by the number of different forms. In
comparison, JavaScript \eval usage was more straightforward and more
predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with
only one form. On the other hand, there are \packageNbSimpleMinimizedOne sites
that only receive one of the simple forms (\ie\xspace \c{X}, \c{V}, \c{\$} or
\c{<-}). Additional \packageNbSimpleMinimizedMore sites takes multiple forms,
but all of them are these simple ones. Among these sites there are many cases
where \eval could be replaced by simpler constructs such as \c{get} to lookup a
name in an environment, \c{assign} to set a value to a name in an environment
or \c{do.call} to reflectively call a function.

@@

\mypara{Discussion.}
The data presented in this section is what one could expect. For {\it
  functions}, they represent the main user case for \eval. Thus \eval is really
extending the behavior of that function by executing a dynamically selected
piece of code in the function's scope. Typically, variables are read (mostly),
written (less), but there are also some cases where new variables are injected
or existing variables are deleted. Usually, this happens in the current
function, but also in frames arbitrarily far up the call stack. One data point
we lack is how the target environment was obtained. The expected case is that
the body of a promise was modified and the resulting expression was evaluated in
the same environment the promise originated from. Likely less frequent are cases
where \eval is provided the results of programmatically selecting some call
stack. For {\it synthetics}, their relatively high frequency correlates to use
cases where one wants to evaluate an expression in either a restricted
environment or to use a data structure as an environment. For {\it global}, its
relatively high frequency is likely an artifact of how the code of our corpus is
is run. For {\it packages}, only \packageNbPackageNamespaceEnvSites sites have
this kind. This is probably for the best as mutating the bindings of
a loaded package is improper and R tries to make it difficult.

@@

\mypara{Discussion}
The origin data suggests that constructing expressions from strings is a
minority of the use cases. Instead, the constructed category shows that most
\evals comes from code that was processed by the compiler, and may be slightly
modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those. Another use of \eval is to directly build symbols to evaluate them in another environment than the default one.
